Death by Dropdown? Use Behavioral Science To Engineer Insightful Shiny Apps

### 2025-09-03 conference-materials/2025_posit-conf/draft.qmd
Streamlines presentation content and replaces external images
- Removes verbose code output examples and telemetry details to focus on core concepts
- Replaces external image links with local files for better reliability and removes unnecessary captions
- Updates GIF embeds to use Tenor widgets for improved loading and accessibility

::: {.fragment}
![](https://media0.giphy.com/media/v1.Y2lkPTc5MGI3NjExNGMxamprcTF2OWgzc2YxeXp0dndlbHNkdHN0NjJhd2JwcWJsbXJncCZlcD12MV9pbnRlcm5hbF9naWZfYnlfaWQmY3Q9Zw/kHU8W94VS329y/giphy.gif)
:::

@@

::: {.fragment}
![](https://i.redd.it/h7t0kz1234567.jpg)<br>
*"But it has ALL the features!"*
:::

@@

```{r}
#| eval: false
library(bidux)

# Define your destination and purpose
interpret_stage <- bid_interpret(
  central_question = "Are we on track this quarter?",
  data_story = list(
    hook = "Q3 performance exceeds targets by 15%",
    context = "Following our new product launch",
    tension = "However, churn increased 8% in enterprise segment",
    resolution = "Focus retention efforts on high-value accounts"
  ),
  user_personas = list(
    primary = "Product managers checking weekly KPIs",
    secondary = "Executives reviewing quarterly trends"
  )
)

# Auto-generates checklist and suggestions
print(interpret_stage)
#> ✓ Central question defined
#> ✓ Data story structured with clear arc
#> ✓ User personas identified
#> → Suggestion: Consider adding success metrics
```

@@

## The Developer's Dilemma

![When the client asks for "just one more filter"](https://i.imgflip.com/8z456x.jpg)

@@

## NOTICE: With {bidux} + Telemetry

```{r}
#| eval: false
# Real telemetry data reveals hidden friction
notice_stage <- bid_notice(
  previous_stage = interpret_stage,
  problem = "Analysis paralysis from filter overload",
  evidence = list(
    quantitative = "65% abandonment rate; 4.3 min to first click",
    qualitative = "Users report feeling overwhelmed",
    telemetry = "Heatmap shows 80% ignore bottom filters"
  ),
  behavioral_theory = "choice_overload"  # Auto-suggests if omitted
)

# Telemetry-driven insights
notice_stage$get_telemetry_insights()
#> 🔍 Telemetry Analysis:
#> • Hover patterns show confusion over filter interactions
#> • Click sequences reveal users reset filters 3x on average
#> • Suggestion: Implement smart defaults based on user role
```

@@

```{r}
#| eval: false
anticipate_stage <- bid_anticipate(
  previous_stage = notice_stage,
  include_accessibility = TRUE
)

# Real-time bias detection
anticipate_stage$add_telemetry_rule(
  trigger = "user_hovers_baseline_3_seconds",
  action = "surface_reference_annotation",
  rationale = "Prevent slope misinterpretation"
)

# Example output from production dashboard
#> 📊 Telemetry Alert (2024-03-15):
#> • Pattern detected: Users consistently misread trend direction
#> • Auto-intervention: Added directional arrows + % change labels
#> • Result: 47% reduction in misinterpretation (n=234 sessions)
```

@@

![](https://i.imgflip.com/8z457m.jpg)

@@

```{r}
#| eval: false
structure_stage <- bid_structure(previous_stage = anticipate_stage)

# Generate Shiny UI code with behavioral principles baked in
structure_stage$generate_ui_scaffold()
#> Generated UI with:
#> ✓ F-pattern scanning optimization
#> ✓ 7±2 chunking for controls
#> ✓ Progressive disclosure triggers
#> ✓ Gestalt grouping principles

# Real production example
structure_stage$apply_telemetry_layout(
  heatmap_data = user_interaction_logs,
  optimize_for = "first_time_users"
)
#> 🎯 Layout optimized based on 1,247 user sessions
#> • Moved key metric to position (100, 50)
#> • Grouped filters by 73% co-selection rate
#> • Hidden advanced options (used by <5% users)
```

@@

![The moment of dashboard enlightenment](https://media.giphy.com/media/OK27wINdQS5YQ/giphy.gif)

@@

summary(validate_stage)

# Production telemetry example
#> 📈 Session Analysis (User #4821):
#> • Journey time: 2.3 minutes (↓ from 4.3 min average)
#> • Insights discovered: 3 key findings
#> • Actions taken: Exported report, shared with team
#>
#> Generated Summary:
#> "Churn in enterprise segment requires immediate attention.
#>  Recommended actions:
#>  1. Schedule retention calls with top 10 at-risk accounts
#>  2. Adjust pricing for mid-tier enterprise plans
#>  3. Review Q4 product roadmap for enterprise features"
```

@@

![](img/before-overload.png)
- 18 filters, 6 tabs
- No clear path
- 65% abandonment
- 4.3 min to insight
:::

@@

![](img/after-complete.png)
- 3 smart filters
- Guided journey
- 85% completion
- 45 sec to insight
:::
::::

@@

```{r}
#| eval: false
# Actual telemetry from bidux-powered dashboards
library(bidux)

bid_telemetry_report(
  app_id = "quarterly_kpi_dashboard",
  period = "2024_Q4"
)
```

@@

::: {.column width="40%"}
::: {.fragment}
![Write it down, stick it on your monitor]()
:::
:::
::::

@@

::: {.incremental}
1. **Download {bidux}** → `install.packages("bidux")`
2. **Try ONE stage** on your worst dashboard → `bid_quickstart("interpret")`
3. **Connect with me** to share your before/after → [@jrwinget](https://bsky.app/profile/jrwinget.bsky.social)
:::

::: {.fragment}
**Bonus:** First 10 people to share their transformation get free 1:1 dashboard review! 🎁
:::

@@

::: {.notes}
- Remind about mnemonic: "Interfaces Need A Smart Vision"
- Mention I'll be at the Shiny booth with laptop for live demos
- Thank organizers
- Quick reminder about the free dashboard reviews
:::

>>>>

::: {.fragment}
<div class="tenor-gif-embed" data-postid="17038619" data-share-method="host" data-aspect-ratio="1.77778" data-width="100%"><a href="https://tenor.com/view/ron-swanson-parks-and-recreation-parks-and-rec-computer-dumpster-gif-17038619">Ron Swanson Parks And Recreation GIF</a>from <a href="https://tenor.com/search/ron+swanson-gifs">Ron Swanson GIFs</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::

@@

::: {.fragment}
![](img/this-is-fine.jpg)<br>
*"But it has ALL the features!"*
:::

@@

```{r}
#| eval: false

library(bidux)

# Define your destination and purpose
interpret_stage <- bid_interpret(
  central_question = "Are we on track this quarter?",
  data_story = list(
    hook = "Q3 performance exceeds targets by 15%",
    context = "Following our new product launch",
    tension = "However, churn increased 8% in enterprise segment",
    resolution = "Focus retention efforts on high-value accounts"
  ),
  user_personas = list(
    primary = "Product managers checking weekly KPIs",
    secondary = "Executives reviewing quarterly trends"
  )
)
```

@@

## When the client asks for "just one more filter"

![](img/anakin.jpg)

@@

## NOTICE: With {bidux}

```{r}
#| eval: false

# Real telemetry data reveals hidden friction
notice_stage <- bid_notice(
  previous_stage = interpret_stage,
  problem = "Analysis paralysis from filter overload",
  evidence = list(
    quantitative = "65% abandonment rate; 4.3 min to first click",
    qualitative = "Users report feeling overwhelmed",
    telemetry = "Heatmap shows 80% ignore bottom filters"
  )
)
```

@@

```{r}
#| eval: false

anticipate_stage <- bid_anticipate(
  previous_stage = notice_stage,
  include_accessibility = TRUE
)
```

@@

![](img/oprah.jpg)

@@

```{r}
#| eval: false

structure_stage <- bid_structure(previous_stage = anticipate_stage)
```

@@

![](https://media.giphy.com/media/OK27wINdQS5YQ/giphy.gif)

@@

summary(validate_stage)
```

@@

- 18 filters, 6 tabs
- No clear path
- 65% abandonment
- 4.3 min to insight
:::

@@

![](img/after-complete.png)

- 3 smart filters
- Guided journey
- 85% completion
- 45 sec to insight
:::
::::

@@

```{r}
#| eval: false

# Actual telemetry from bidux-powered dashboards
bid_report(
  app_id = "quarterly_kpi_dashboard",
  period = "2024_Q4"
)
```

@@

::: {.column width="40%"}
::: {.fragment}
<div class="tenor-gif-embed" data-postid="5313946" data-share-method="host" data-aspect-ratio="1.3" data-width="100%"><a href="https://tenor.com/view/writethatdown-gif-5313946">Writethatdown GIF</a>from <a href="https://tenor.com/search/writethatdown-gifs">Writethatdown GIFs</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::
:::
::::

::: {.notes}
- for gif: "Write that down! Stick it to your monitor!"
:::

@@

::: {.incremental}
1. **Download {bidux}** → `install.packages("bidux")`
2. **Try a some of the auto-suggestions** on a completed/WIP dashboard
3. **Connect with me** to share your before/after → [@jrwinget](https://bsky.app/profile/jrwinget.bsky.social)
:::

::: {.fragment}
**Bonus:** First 5 people to share their transformation get free 1:1 dashboard review! 🎁
:::

@@

::: {.notes}
- Remind about mnemonic: "Interfaces Need A Smart Vision"
- Feel free to reach out if you'd like to connect and chat further
- Thank organizers
- Quick reminder about the free dashboard reviews
:::
